Financialisation of commodities: co-movement behind-the-scenes

### 2022-05-26 docs/paper.tex
revision first pass

  \begin{abstract}
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon--commonly referred to as ``financialisation''---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financilisation that comprehends commodity futures as an asset class. Using a bespoke asset pricing factors based framework we study the issue through the lens of co-movement and find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals, in other words, a process of ``definancialisation'' in global commodity futures markets.
  \end{abstract}

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialisation\footnote{Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits (Büyükşahin and Robe, 2014; Singleton, 2013), institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialisation of commodities (Domanski and Heath, 2007).} as well as the period of the 2008 financial crisis and its aftermath. Financialisation appears to have affected multiple economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A vigorous legislative debate accompanied by an equally intense policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors\footnote{See section \ref{co-movement-background} for a detailed discussion.} eventually led to regulatory changes in several individual markets.\\
In response to immediate policy concerns, the academic debate was initially framed rather narrowly around adequacy of speculation--the burning issue of the day--at the individual commodity market level and the ensuing academic analysis focused on the more mechanical effects of financialisation. With the benefit of hindsight, this approach seems to lack a fundamental understanding of the phenomenon it endeavoured to study and which consequences it attempted to address.\\
Financialisation was a phenomenon global in nature; through the channel of index investment it swayed the whole cross-section of liquid commodities at once, eventually spurring its transition away from a collection of idiosyncratic markets to its emergence as an asset class. Focusing on individual markets in this context thus seems incomprehensive. The mechanical approach similarly seems amiss both in form and substance. Most of the studies underpinning the regulatory response to financialisation took the form of pairwise correlation, causation and to some extent co-integration analysis, all of which are idiosyncratic in nature for they study single pairs of individual assets. Besides, some of these techniques since proved controversial and seem to produce results the quality of which may fall short of the standards that may reasonably be expected to lay the ground for legislative action. In substance, leaving the idiosyncrasy issue aside, a case could be made for the mechanical approach under the assumption that financialised commodity markets can be described, ceteris paribus, as their prior or legacy state with increased uninformed financial participation. The aforementioned technique(s) could thence be used to observe potential changes in pricing dynamics, for example, which origin could be naturally attributed to the market behaviour of the new comers. Although appealing, for it comes with the promise of prompt ``significant'' results, this approach misses a critical corollary of the substantial advent of a new category of players in a (set of) market(s): the alteration of market fundamentals it potentially induces, best observed in this context through the lens of commercial hedging pressure (Hicks, 1939; Keynes, 1930).\\
While essential to the understanding of the phenomenon, the global nature of financialisation and its impact on market fundamentals therefore appear to have been originally overlooked and another approach, perhaps more refined, seems warranted.

@@

Financialisation was an issue of such policy importance that it triggered legislative action. The debate was initially framed around adequacy of speculation, the burning issue of the day, and the ensuing academic analysis focused on the more mechanical effects of financialisation.\\
With the benefit of hindsight, it appears this approach was perhaps too narrow, and it now seems necessary to address the phenomenon from a broader perspective. Commodity price dynamics appear to have altered substantially in quite different ways over the financialisation and the financial crisis periods and our commodity futures asset pricing factors based approach seems able to provide new insights into the nature of these changes: financialisation was a phenomenon endogenous to the commodity markets transmitted via the commodity futures markets and which effects were particularly strong for the metals sector while the crisis and its aftermath seems to have delivered an exogenous shock across the entire cross-section of liquid commodities and thereby potentially triggered a process of ``definancialisaion'' in global commodity futures markets.\\
The initial view of financialisation was that it consisted of speculative flows which had the effect of driving up and creating bubble like conditions in commodity futures prices\footnote{Masters (2008); Masters and White (2008); UNCTAD and Cooperation (2009); De Schutter (2010); Gilbert (2010a); Gilbert (2010b); Herman et al. (2011); Schumann (2011); Singleton (2013).}. However, the fundamental question of the nature of the impact of financialisation across the entire cross-section of commodity futures markets has not yet been completely answered. We provide an empirical complement to a new stream of theoretical studies that try to model the impact of financialisation on various aspects of commodity futures markets\footnote{Etula (2013), Acharya et al. (2013), Cheng et al. (2014), Leclercq and Praz (2014), Sockin and Xiong (2015), Goldstein et al. (2014), Ekeland et al. (2016), Goldstein and Yang (2017).} and demonstrate that the effects of financialisation extend beyond the mechanical effects induced by indexation as outlined in Tang and Xiong (2012) and Basak and Pavlova (2016).

@@

Using our bespoke commodity-based asset pricing framework we strive to shed new lights on the phenomenon by first regressing returns series for the twenty four individual US commodities against our factor mimicking portfolios independently. The average time series explanatory power of the various one factor models on the US traded assets over the four periods is shown in Table \ref{tab:co-movement-time-reg-US-facts-US}.\\
Over the first period the market factor has the best average time series performance with the highest average \(R^{2}\) with all three futures market factors performing poorly and the overall performance consistent across periods of high and low \textbf{CHP} (below period median). The two factor models average \(R^{2}\)s (results available on the \href{http://18.135.131.217:3838/financialization-asset-pricing/}{online results dashboard}) are roughly the sum of the one factor \(R^{2}\)s indicating that the factors are almost orthogonal over this period.\\
The onset of financialisation leads to a substantial improvement in the performance of the CHP factor with its average \(R^{2}\) increasing to 8.2\% in the second period relative to 1.7\% in the first with this improvement coming from the long leg that shows a greater improvement (4.5\% to 15.1\%) relative to the short (3\% to 4\%) bringing more evidence of the important role played by long only investors in the change in nature of these markets over that period. The performance of the market factor also improves though not by as much in a relative sense (21.6\% from 11.7\%) and both the term structure factor and the open interest growth factor show a much smaller improvement in performance with smaller differences between legs. Over all periods, all of our factor models show similar performance across regimes of high and low \textbf{CHP}.

@@

In table \ref{tab:co-movement-correlations}, we show the average pairwise returns correlation between each set of picks in each time period. This exercise gives an indication of the extent to which attribute co-movement reveals price co-movement, which is particularly relevant for the long-short factors for which the corresponding attribute is not directly related to price co-movement.\\
In the first period the average picks pairwise correlation for the long-short factors ranges from 7.1\% (term structure factor picks) to 12.4\% (open interest factor picks) while the market factor picks show the highest figure (23.7\%). In the second period the figures for the long-short factors rise sharply; up to 22.9\% from 8.1\% for the average CHP factor picks, up to 32.4\% from 12.4\% for the open interest factor picks and up to 23.4\% from 7.1\% for the term structure factor picks. This increase is stronger than that for the market factor picks in a relative sense and is a consequence of the dramatic increase in average pairwise price correlation amongst the US traded metals which dominate the average CHP factor set of picks and which presence is stronger amongst both the open interest and term structure factor sets of picks over the financialisation period. The higher figures for the open interest factor are driven by the strong presence of the energy complex amongst the corresponding picks. The latter exhibits much higher average pairwise correlations than the other sectors in both periods although lower in the second relative to the first which reinforces the role played by the metals in the sharp increase in average pairwise correlation amongst picks for this factor.\\
With regards to \textbf{CHP} regimes, although not very pronounced, a pattern is noticeable. For the market, open interest and term structure factor picks, average pairwise correlations are stronger over phases of \textbf{contango} in the first period and over phases of \textbf{backwardation} in the financialisation period, eventually reverting back to being stronger over \textbf{contango} phases in the crisis period with the CHP factor sets of picks showing the opposite pattern. Overall, these results suggest an increase in correlation between non-price attributes co-movement and price co-movement over the financialisation period. It is also interesting to note that US traded metals returns were negatively correlated with the rest of US traded commodities in the first period and positively in the second.

@@

Taken together the patterns of results over the crisis and post-crisis periods echo those in Basu and Bauthéac (2021) in suggesting that the exogenous shock delivered by the financial crisis combined with the accommodative monetary policy that ensued seem to have initiated a reversion toward pre-financialisation fundamentals in the commodity futures complex, in other words, a process of ``definancialisation''.

@@

With the benefit of hindsight, this focus appears to have been too narrow. As our results suggests, financialisation was a phenomenon global in nature, with cross-sectional effects on market fundamentals and studying the direct effects of increased market participation in single markets thus seems amiss in this context.\\
The results of our analysis further suggest that financialisation set off an ontological drift in the commodity complex beyond energy and agricultural commodities and in fact strongest for the metals sector at the global level; an aspect largely overlooked in the original approach that we believe the existing literature has not fully analysed. The regulatory changes underpinned by these studies may thus have been unwarranted with potential substantial undesired effects on the functioning of the markets in which they were enforced and thence, to a larger extent on that of the economic system as a whole.\\
The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, also appear to have induced significant changes in commodity pricing dynamics. In contrast with financialisation, our results suggest that the crisis and its aftermath have delivered an exogenous shock across the entire cross-section of liquid commodities and seem to have triggered a motion of reversion to legacy pre-financialisation fundamentals, in other words, a process of ``definancialisation'' in global commodity futures markets.

>>>>

  \begin{abstract}
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon--commonly referred to as ``financialisation''---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financilisation that comprehends commodity futures as an asset class. Using a bespoke asset pricing factors based framework we study the issue through the lens of co-movement and find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
  \end{abstract}

@@

The last twenty years have seen major upheavals in the commodity markets with the onset of financialisation\footnote{Starting around 2004, with a view of commodity markets as an emerging asset class and in an effort to hedge against inflation and seek diversification benefits (Büyükşahin and Robe, 2014; Singleton, 2013), institutional investors sent forth an unprecedented flow of capital into commodity futures markets. This phenomenon is commonly referred to as the financialisation of commodities (Domanski and Heath, 2007).} as well as the period of the 2008 financial crisis and its aftermath. Financialisation appears to have affected multiple economic sectors, including agriculture and energy, and its impact has been hotly debated in the policy, legislative, regulatory, and academic spheres. A vigorous legislative debate accompanied by an equally intense policy and academic debate across a number of disciplines about the perceived effects on commodity prices of the unprecedented inflow of institutional funds brought into commodity futures markets by new index type investors\footnote{See section \ref{co-movement-background} for a detailed discussion.} eventually led to regulatory changes in several individual markets.\\
In response to immediate policy concerns, the academic debate was initially framed rather narrowly around adequacy of speculation--the burning issue of the day--at the individual commodity market level and the ensuing academic analysis focused on the more mechanical effects of financialisation. With the benefit of hindsight, this approach seems to lack a fundamental understanding of the phenomenon it endeavoured to study and which consequences it attempted to address.\\
Financialisation was a phenomenon global in nature; through the channel of index investment it affected the whole cross-section of liquid commodities at once, eventually spurring its transition away from a collection of idiosyncratic markets to its emergence as an asset class. Focusing on individual markets in this context thus seems limited. Most of the studies underpinning the regulatory response to financialisation took the form of pairwise correlation, causation and to some extent co-integration analysis, all of which are idiosyncratic in nature for they study single pairs of individual assets. Besides, some of these techniques since proved controversial and seem to produce results the quality of which may fall short of the standards that may reasonably be expected to lay the ground for legislative action. Although appealing, for it comes with the promise of prompt ``significant'' results, this approach misses a critical corollary of the substantial advent of a new category of players in a (set of) market(s): the alteration of market fundamentals it potentially induces, best observed in this context through the lens of commercial hedging pressure (Hicks, 1939; Keynes, 1930).\\
While essential to the understanding of the phenomenon, the global nature of financialisation and its impact on market fundamentals therefore appear to have been originally overlooked and another approach, perhaps more refined, seems warranted.

@@

Financialisation was an issue of such policy importance that it triggered legislative action. The debate was initially framed around adequacy of speculation, the burning issue of the day, and the ensuing academic analysis focused on the more mechanical effects of financialisation.\\
With the benefit of hindsight, it appears this approach was perhaps too narrow, and it now seems necessary to address the phenomenon from a broader perspective. Commodity price dynamics appear to have altered substantially in quite different ways over the financialisation and the financial crisis periods and our commodity futures asset pricing factors based approach seems able to provide new insights into the nature of these changes: financialisation was a phenomenon endogenous to the commodity markets transmitted via the commodity futures markets and which effects were particularly strong for the metals sector while the crisis and its aftermath seems to have delivered an exogenous shock across the entire cross-section of liquid commodities and thereby potentially triggering a reversion to pre-financialisation dynamics.

The initial view of financialisation was that it consisted of speculative flows which had the effect of driving up and creating bubble like conditions in commodity futures prices\footnote{Masters (2008); Masters and White (2008); UNCTAD and Cooperation (2009); De Schutter (2010); Gilbert (2010a); Gilbert (2010b); Herman et al. (2011); Schumann (2011); Singleton (2013).}. However, the fundamental question of the nature of the impact of financialisation across the entire cross-section of commodity futures markets has not yet been completely answered. We provide an empirical complement to a new stream of theoretical studies that try to model the impact of financialisation on various aspects of commodity futures markets\footnote{Etula (2013), Acharya et al. (2013), Cheng et al. (2014), Leclercq and Praz (2014), Sockin and Xiong (2015), Goldstein et al. (2014), Ekeland et al. (2016), Goldstein and Yang (2017).} and demonstrate that the effects of financialisation extend beyond the mechanical effects induced by indexation as outlined in Tang and Xiong (2012) and Basak and Pavlova (2016).

@@

Using our bespoke commodity-based asset pricing framework we strive to shed new light on the phenomenon by first regressing returns series for the twenty four individual US commodities against our factor mimicking portfolios independently. The resulting individual explanatory powers (\(R^2\)s), once averaged by factor and period, offer a refined perspective on the relation between return and factor attribute co-movement. A low figure indicates low cross-sectional return co-movement for, in this context, return on the corresponding factor mimicking portfolio only has explanatory power on return for commodities that are constituents of one of its legs over the corresponding period. The explanatory power for the other commodities is typically low or even 0 for the case where commodities are perfectly uncorrelated with one-another. A rising figure on the other hand thus indicates increased pairwise correlations, positive or negative, including amongst factor leg mimicking portfolio constituent commodities, which suggests that pairwise correlation relate to factor attribute.\\
The average time series explanatory power of the various one factor models on the US traded assets over the four periods is shown in Table \ref{tab:co-movement-time-reg-US-facts-US}. Over the first period the market factor has the best average time series performance with the highest average \(R^{2}\) with all three futures market factors performing poorly, indicating that the commodities with extreme factor attributes (high or low CHP for example) are uncorrelated with the rest of the commodity cross-section, and the overall performance consistent across periods of high and low \textbf{CHP} (below period median). The two factor models average \(R^{2}\)s (results available on the \href{http://18.135.131.217:3838/financialization-asset-pricing/}{online results dashboard}) are roughly the sum of the one factor \(R^{2}\)s indicating that the factors are almost orthogonal over this period.\\
The onset of financialisation leads to a substantial improvement in the performance of the CHP factor with its average \(R^{2}\) increasing to 8.2\% in the second period relative to 1.7\% in the first with this improvement coming from the long leg that shows a greater improvement (4.5\% to 15.1\%) relative to the short (3\% to 4\%), suggesting that ``normally'' backwardated commodities (Keynesian perspective) have higher correlation with the rest of the commodity cross-section, and thereby bringing more evidence of the important role played by long only investors in the change in nature of these markets over that period. The performance of the market factor also improves though not by as much in a relative sense (21.6\% from 11.7\%) and both the term structure factor and the open interest growth factor show a much smaller improvement in performance with smaller differences between legs. Over all periods, all of our factor models show similar performance across regimes of high and low \textbf{CHP}.

@@

In table \ref{tab:co-movement-correlations}, we show the average pairwise returns correlation between each set of picks in each time period. Following on form the earlier observations, this exercise gives an indication of the extent to which attribute co-movement reveals price co-movement, which is particularly relevant for the long-short factors for which the corresponding attribute is not directly related to price co-movement.\\
In the first period the average picks pairwise correlation for the long-short factors ranges from 7.1\% (term structure factor picks) to 12.4\% (open interest factor picks) while the market factor picks show the highest figure (23.7\%). In the second period the figures for the long-short factors rise sharply; up to 22.9\% from 8.1\% for the average CHP factor picks, up to 32.4\% from 12.4\% for the open interest factor picks and up to 23.4\% from 7.1\% for the term structure factor picks. This increase is stronger than that for the market factor picks in a relative sense and is a consequence of the dramatic increase in average pairwise price correlation amongst the US traded metals which dominate the average CHP factor set of picks and which presence is stronger amongst both the open interest and term structure factor sets of picks over the financialisation period. The higher figures for the open interest factor are driven by the strong presence of the energy complex amongst the corresponding picks. The latter exhibits much higher average pairwise correlations than the other sectors in both periods although lower in the second relative to the first which reinforces the role played by the metals in the sharp increase in average pairwise correlation amongst picks for this factor.\\
With regards to \textbf{CHP} regimes, although not very pronounced, a pattern is noticeable. For the market, open interest and term structure factor picks, average pairwise correlations are stronger over phases of \textbf{contango} in the first period and over phases of \textbf{backwardation} in the financialisation period, eventually reverting back to being stronger over \textbf{contango} phases in the crisis period with the CHP factor sets of picks showing the opposite pattern. Overall, these results suggest an increase in correlation between non-price attributes co-movement and price co-movement over the financialisation period. It is also interesting to note that US traded metals returns were negatively correlated with the rest of US traded commodities in the first period and positively in the second.

@@

Taken together the patterns of results over the crisis and post-crisis periods echo those in Basu and Bauthéac (2021) in suggesting that the exogenous shock delivered by the financial crisis combined with the accommodative monetary policy that ensued seem to have initiated a reversion toward pre-financialisation fundamentals in the commodity futures complex.

@@

With the benefit of hindsight, this focus appears to have been too narrow. As our results suggests, financialisation was a phenomenon global in nature and studying the direct effects of increased market participation in single markets thus seems amiss in this context.\\
The results of our analysis further suggest that financialisation had cross-sectional effects on market fundamentals across the commodity complex beyond energy and agricultural commodities and in fact strongest for the metals sector at the global level; an aspect largely overlooked in the original approach that we believe the existing literature has not fully analysed.\\
The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, also appear to have induced significant changes in commodity pricing dynamics. In contrast with financialisation, our results suggest that the crisis and its aftermath have delivered an exogenous shock across the entire cross-section of liquid commodities and seem to have triggered a motion of reversion to legacy pre-financialisation fundamentals.
